---
title: "Annotation Methods"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Annotation Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

CellVoteR runs four primary annotation methods and two global tie-breakers.

# Run CellVoteR

```{r run}
results <- run_cellvoter(sce)
```

# Primary Methods

| Method | Feature space | Broad strategy | Fine annotation |
|---|---|---|---|
| Method 1 | Full HVG space | Cluster-based | Subcluster marker scoring |
| Method 2 | Reduced marker space | Cluster-based | Subcluster marker scoring |
| Method 3 | Full HVG space | Enrichment-based | Subcluster marker scoring |
| Method 4 | Reduced marker space | Enrichment-based | Subcluster marker scoring |

# Global Tie-breakers

| Tie-breaker | Feature space | Strategy |
|---|---|---|
| Tie-breaker 1 | Full HVG space | Global clustering and marker scoring |
| Tie-breaker 2 | Reduced marker space | Global clustering and marker scoring |

# Fine Annotation Flow

```text
broad annotation
↓
subcluster_labels()
↓
rank_cluster_markers()
↓
extract_top_markers()
↓
score_markers_against_panel()
↓
assign_fine_labels()
```

# Custom Parameters

```{r run-custom}
results <- run_cellvoter(
  sce,
  return_full_output = TRUE,
  annotation_args = list(
    broad_args = list(
      test_type = "wilcox",
      min_prop  = 0.1
    ),
    rank_args = list(
      test_type = "wilcox",
      min_prop  = 0.25
    ),
    extract_args = list(
      fdr_threshold    = 0.05,
      effect_threshold = 0.6,
      target_n         = 100L
    )
  )
)
```

# Full Outputs

```{r full-output}
results$full_output$method_1$scores
results$full_output$global_2$scores
```
