---
title: "CellVoteR: Ensemble Cell Type Annotation for Single-Cell RNA-seq"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{CellVoteR: Ensemble Cell Type Annotation for Single-Cell RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## Overview

CellVoteR is an ensemble-based pipeline for robust cell type annotation in
single-cell RNA-seq (scRNA-seq) data. Rather than relying on a single
classification strategy, CellVoteR integrates four complementary annotation
methods across two feature spaces, then resolves disagreements through a
principled consensus voting step.

The core design philosophy is:

- **Divide and conquer** — broadly triage cells into lineages before applying
  fine-resolution annotation, preventing dominant populations from masking rare
  cell types
- **Redundancy** — four methods running in parallel reduces sensitivity to the
  failure modes of any single approach
- **Separation of concerns** — annotation (slow, compute-intensive) and
  consensus resolution (fast, parameter-sensitive) are decoupled, so the user
  can re-tune voting without repeating the pipeline

### What CellVoteR requires

Two inputs are needed before running the pipeline:

1. A **raw gene-by-cell counts matrix** (sparse `dgCMatrix`, RDS file, or MTX
   triplet)
2. A **marker configuration** — a structured list of broad and fine cell type
   marker genes, loaded from an Excel workbook or nested YAML file

---

## Installation

```{r install}
# Install the development version from GitHub
# install.packages("devtools")
devtools::install_github("ajxa/CellVoteR")
```

```{r load}
library(CellVoteR)
```

---

## Step 1 — Preparing Marker Inputs

Markers are the backbone of CellVoteR's annotation strategy. They are
organised into two tiers:

### Broad markers

Broad markers define coarse cell lineages (e.g. *Immune*, *Vasculature*,
*Other*). They must be:

- **Small** — typically 2–5 genes per category
- **Mutually exclusive** — no gene should appear in more than one broad
  category
- **Biologically diagnostic** — genes that robustly delineate lineages even
  in heterogeneous datasets

Broad markers are loaded and then configured with `build_broad_marker_config()`,
which assigns expression thresholds and priority rankings used for tie-breaking
when a cell passes multiple broad categories.

### Fine markers

Fine markers define sub-populations within each broad lineage (e.g. *B cell*,
*T cell*, *NK cell* within *Immune*). They can be larger gene sets and are
used for Fisher's Exact Test scoring during fine annotation. They do not need
to be mutually exclusive.

### Loading markers

Markers are loaded from an Excel workbook (`.xlsx`) or a nested YAML file
using `load_markers()`. The expected structure has two sheets or top-level
keys: `broad` and `fine`.

```{r load-markers}
markers <- load_markers(file_path = "path/to/input_markers.xlsx")

# Inspect the structure
str(markers$broad)   # named list of character vectors
str(markers$fine)    # nested named list: broad category > fine cell type > genes
```

### Configuring broad markers

`build_broad_marker_config()` processes the raw broad marker list, attaching
expression thresholds and priority ranks used during the enrichment-based
annotation step.

```{r build-broad}
markers$broad <- build_broad_marker_config(
  marker_list       = markers$broad,
  priority_order    = c("vasculature", "immune"),  # higher priority listed first
  default_threshold = 0.25                          # default logcounts threshold
)

# Each broad category now has markers, expr_threshold, coexp_min, and priority
str(markers$broad$immune)
```

The `priority_order` argument controls tie-breaking when a cell passes
expression thresholds for more than one broad category — categories listed
earlier receive a lower (higher priority) numeric rank.

---

## Step 2 — Creating the SingleCellExperiment

CellVoteR works natively with
[`SingleCellExperiment`](https://bioconductor.org/packages/SingleCellExperiment/)
objects. Use `create_sce()` to construct one from your raw data.

### From an in-memory sparse matrix

```{r create-sce-memory}
sce <- create_sce(
  counts        = my_sparse_matrix,  # dgCMatrix, genes x cells
  cell_metadata = my_metadata_df     # data.frame, one row per cell (optional)
)
```

### From file paths

`create_sce()` also accepts file paths, which is useful for large datasets
where the matrix is stored on disk:

```{r create-sce-files}
# From RDS files
sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"   # also accepts .csv or .tsv
)

# From MTX triplet files
sce <- create_sce(
  mtx_file   = "path/to/matrix.mtx.gz",
  cells_file = "path/to/barcodes.tsv",
  genes_file = "path/to/features.tsv"
)
```

---

## Step 3 — Quality Control

`assess_cell_quality()` calculates per-cell QC metrics and optionally removes
low-quality cells before downstream analysis.

```{r qc}
sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)
```

Cells that fail QC are flagged in `colData(sce)$QC_PASS`. Setting
`remove_failed_cells = TRUE` subsets the object to passing cells only.

---

## Step 4 — Normalisation

CellVoteR uses a pooling-based normalisation strategy
([Lun et al. 2016](https://doi.org/10.1186/s13059-016-0947-7)) via
`scran::computePooledFactors`, followed by log-normalisation with
`scuttle::logNormCounts`.

```{r normalise}
sce <- normalize_counts(sce)
```

This adds a `logcounts` assay to the SCE and sets `sizeFactors()`. The
`logcounts` assay is required by all downstream steps.

---

## Step 5 — Building Analysis Tracks

`prepare_sce()` is the key preprocessing step. It:

1. Validates broad and fine marker configurations against the expression matrix
2. Builds two independent feature spaces — the **full HVG space** and the
   **reduced marker-defined space**
3. Runs PCA and unsupervised clustering (Leiden via SNN graph) on each space
4. Attaches the marker configuration and filtered fine markers to the SCE
   metadata
5. Stores the reduced feature space as an `altExp` named `"user_panel"`

```{r prepare}
sce <- prepare_sce(sce, markers)
```

After this step, the SCE has the following structure:

```
sce
├── assays: counts, logcounts
├── rowSubset("broad_hvgs")           ← HVGs used for broad clustering
├── reducedDim("PCA_broad_hvg")       ← PCA on full HVG space
├── colData$cluster_broad_hvg         ← Leiden clusters (full space)
├── colData$cluster_broad_reduced     ← Leiden clusters (reduced space)
├── metadata$marker_config            ← full marker configuration
├── metadata$filtered_fine_markers    ← fine markers present in data
├── metadata$missing_by_label         ← per-label missing marker report
└── altExp("user_panel")              ← reduced feature SCE
    ├── assays: counts, logcounts
    ├── reducedDim("PCA")
    ├── colData$cluster
    ├── metadata$marker_config
    ├── metadata$filtered_fine
    └── metadata$params
```

### Automatic parameter estimation

Clustering parameters (number of PCs, SNN neighbourhood size $k$, Leiden
resolution) are estimated automatically from cell count using bounded
log/sqrt scaling via `estimate_cluster_params()`. You can override these if
needed:

```{r prepare-override}
sce <- prepare_sce(
  sce,
  markers,
  n_hvgs     = 3000L,
  n_pcs      = 30L,
  k          = 20L,
  resolution = 0.8
)
```

### Marker overlap reporting

If fine markers are partially missing from your dataset, `prepare_sce()`
reports which labels are most affected:

```{r missing-markers}
# Inspect missing marker report after prepare_sce()
metadata(sce)$missing_by_label
```

Missing markers are automatically removed from the fine marker sets used for
scoring, so annotation still proceeds with the genes that are present.

---

## Step 6 — Running the Ensemble

`run_cellvoter()` orchestrates all six annotation pipelines and returns a
named list of per-cell label factors, one per method.

```{r run}
results <- run_cellvoter(sce)
```

### What happens internally

Four primary methods and two global tie-breakers are run:

| Method | Feature space | Broad strategy | Subcluster feature mode |
|---|---|---|---|
| Method 1 | Full (HVG) | Cluster-based | HVG |
| Method 2 | Reduced (panel) | Cluster-based | All |
| Method 3 | Full (HVG) | Enrichment-based | HVG |
| Method 4 | Reduced (panel) | Enrichment-based | All |
| Tie-breaker 1 | Full (HVG) | None (global) | — |
| Tie-breaker 2 | Reduced (panel) | None (global) | — |

Each primary method follows the same pipeline:

```
broad annotation
      ↓
subcluster_labels()
      ↓
rank_cluster_markers()    ← DE testing per sub-cluster
      ↓
extract_top_markers()     ← select top N genes per cluster
      ↓
score_markers_against_panel()   ← Fisher's Exact Test + overlap similarity
      ↓
assign_fine_labels()      ← best label per cluster → mapped to cells
```

### Broad annotation strategies

**Cluster-based** (`annotate_broad_clusters`): Runs DE testing on pre-existing
unsupervised clusters. Each cluster is assigned the broad category whose
curated markers have the lowest median rank among significantly upregulated
genes (FDR ≤ 0.05, AUC ≥ 0.6 by default).

**Enrichment-based** (`annotate_broad_cells`): Assigns labels directly to
individual cells by aggregating expression across each broad category's marker
genes and comparing against category-specific thresholds. Does not depend on
clustering.

### Edge case — collapsed broad labels

In some datasets (e.g. a highly homogeneous tumour sample), all clusters may
receive the same broad label. CellVoteR detects this and retains the original
cluster structure for subclustering rather than collapsing to a single group,
preserving fine-resolution information downstream.

### Returned colData columns

After `run_cellvoter()`, the returned `$sce` has all intermediate cluster
labels attached:

| Column | Description |
|---|---|
| `cluster_broad_hvg` | Pre-existing HVG clusters from `prepare_sce()` |
| `cluster_broad_reduced` | Pre-existing reduced clusters from `prepare_sce()` |
| `broad_cluster_m1` | Broad labels, method 1 |
| `broad_cluster_sub_m1` | Sub-cluster labels, method 1 |
| `broad_cluster_m2` | Broad labels, method 2 |
| `broad_cluster_sub_m2` | Sub-cluster labels, method 2 |
| `broad_enrichment_m3` | Broad labels, method 3 |
| `broad_enrichment_sub_m3` | Sub-cluster labels, method 3 |
| `broad_enrichment_m4` | Broad labels, method 4 |
| `broad_enrichment_sub_m4` | Sub-cluster labels, method 4 |

### Customising annotation parameters

`run_cellvoter()` accepts an `annotation_args` list to customise the
underlying functions. There are two independent ranking steps, each with its
own sublist:

```{r run-custom}
results <- run_cellvoter(
  sce,
  return_full_output = TRUE,
  annotation_args = list(

    # Controls ranking inside annotate_broad_clusters (methods 1 and 2)
    # Lenient min_prop appropriate for small, specific broad marker sets
    broad_args = list(
      test_type = "wilcox",
      min_prop  = 0.1
    ),

    # Controls ranking inside .run_fine_annotation (all six methods)
    # Stricter settings for fine sub-cluster marker extraction
    rank_args = list(
      test_type = "wilcox",
      min_prop  = 0.25
    ),

    # Controls top marker extraction
    extract_args = list(
      fdr_threshold    = 0.05,
      effect_threshold = 0.6,
      target_n         = 100L
    )
  )
)
```

> **Note on `broad_args` vs `rank_args`:** Both control
> `rank_cluster_markers()` but at different pipeline stages.
> `broad_args` affects how broad lineage labels are assigned (methods 1 and 2
> only); `rank_args` affects how sub-cluster marker genes are extracted for
> Fisher scoring (all six methods). Methods 3 and 4 use `rank_args` only,
> as the enrichment-based broad step does not call `rank_cluster_markers()`.

### Accessing full outputs

When `return_full_output = TRUE`, per-cluster Fisher scores and similarity
values are available for every method:

```{r full-output}
# Per-cluster score table for method 1
results$full_output$method_1$scores

# Per-cluster score table for tie-breaker 2
results$full_output$global_2$scores
```

---

## Step 7 — Resolving Consensus Labels

Consensus resolution is intentionally a separate step. This means you can
adjust voting parameters and re-run `resolve_consensus_labels()` as many
times as needed without re-running the annotation pipeline.

```{r resolve}
consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = FALSE,
  ordered_tiebreak  = TRUE
)
```

### Decision hierarchy

For each cell, the following logic is applied in order:

1. **Strong majority** — if one label receives strictly more than 50% of
   method votes (3 of 4 by default) it is assigned immediately.

2. **Tie-breaker agreement** — if the two tie-breakers agree with each other
   *and* their shared label matches the leading candidate, that label is
   assigned.

3. **Ordered tie-breaking** (`ordered_tiebreak = TRUE`, default) — tie-breaker
   1 is tried first; if it matches the leading candidate the label is assigned.
   Otherwise tie-breaker 2 is tried.

4. **Either tie-breaking** (`ordered_tiebreak = FALSE`) — either tie-breaker
   agreeing with the leading candidate is sufficient, with no priority between
   them.

5. **Unresolved** — if every method disagrees (no leading candidate exists) or
   no tie-breaker resolves the split, `unassigned_label` is assigned.

### Voting parameters

| Parameter | Default | Effect |
|---|---|---|
| `allow_even_split` | `FALSE` | When `TRUE`, a 2-of-4 plurality is accepted as majority |
| `ordered_tiebreak` | `TRUE` | When `FALSE`, either tie-breaker can resolve a split |
| `unassigned_label` | `"Unknown"` | Label for unresolved cells |

### Attaching labels to the SCE

```{r attach}
sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method
```

---

## Step 8 — Inspecting Results

### Summary tables

```{r inspect}
# Final label distribution
table(sce$cellVoteR_label)

# Decision method breakdown
table(sce$cellVoteR_method)

# Cross-tabulate label vs decision method
table(sce$cellVoteR_label, sce$cellVoteR_method)
```

### Per-method label distributions

Comparing individual method outputs before consensus is useful for
understanding where methods agree and disagree:

```{r per-method}
table(results$labels$method_1)   # cluster-based, full
table(results$labels$method_2)   # cluster-based, reduced
table(results$labels$method_3)   # enrichment-based, full
table(results$labels$method_4)   # enrichment-based, reduced
table(results$labels$global_1)   # tie-breaker 1
table(results$labels$global_2)   # tie-breaker 2
```

### Method agreement

A useful diagnostic is to look at how often all four primary methods agree:

```{r agreement}
label_df <- as.data.frame(results$labels[1:4])

# Proportion of cells where all four methods agree
mean(apply(label_df, 1, function(x) length(unique(x)) == 1L))

# Per-cell method agreement count
label_df$n_agree <- apply(label_df, 1, function(x) max(table(x)))
table(label_df$n_agree)
```

### Re-running consensus with different parameters

```{r re-resolve}
# Try allowing even splits
consensus_liberal <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = TRUE
)

table(consensus_liberal$label)

# Compare unresolved rate between settings
mean(consensus$label == "unknown")
mean(consensus_liberal$label == "unknown")
```

---

## Complete Workflow

The full pipeline from raw data to annotated SCE:

```{r full-workflow}
library(CellVoteR)

# ── 1. Load and configure markers ──────────────────────────────────────────────
markers <- load_markers(file_path = "path/to/input_markers.xlsx")

markers$broad <- build_broad_marker_config(
  marker_list       = markers$broad,
  priority_order    = c("vasculature", "immune"),
  default_threshold = 0.25
)

# ── 2. Create SCE ──────────────────────────────────────────────────────────────
sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"
)

# ── 3. QC ──────────────────────────────────────────────────────────────────────
sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)

# ── 4. Normalise ───────────────────────────────────────────────────────────────
sce <- normalize_counts(sce)

# ── 5. Build analysis tracks ───────────────────────────────────────────────────
sce <- prepare_sce(sce, markers)

# ── 6. Run ensemble annotation ─────────────────────────────────────────────────
results <- run_cellvoter(sce)

# ── 7. Resolve consensus ───────────────────────────────────────────────────────
consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown"
)

# ── 8. Attach labels ───────────────────────────────────────────────────────────
sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method

# ── 9. Inspect ─────────────────────────────────────────────────────────────────
table(sce$cellVoteR_label)
table(sce$cellVoteR_method)
```

---

## Marker File Format

### Excel workbook

The expected workbook has two sheets:

**Sheet: `broad`**

| category | gene |
|---|---|
| immune | CD45 |
| immune | PTPRC |
| vasculature | PECAM1 |
| vasculature | CDH5 |

**Sheet: `fine`**

| broad_category | cell_type | gene |
|---|---|---|
| immune | B cell | CD79A |
| immune | B cell | CD79B |
| immune | T cell | CD3D |
| immune | T cell | IL32 |
| vasculature | Endothelial | CLDN5 |

### YAML format

```yaml
broad:
  immune:
    - CD45
    - PTPRC
  vasculature:
    - PECAM1
    - CDH5

fine:
  immune:
    B cell:
      - CD79A
      - CD79B
    T cell:
      - CD3D
      - IL32
  vasculature:
    Endothelial:
      - CLDN5
      - SPARC
```

---

## Tips and Troubleshooting

### Low marker overlap

If `prepare_sce()` warns about low fine marker overlap, inspect which labels
are most affected:

```{r missing}
metadata(sce)$missing_by_label
```

Consider whether the missing genes are platform-specific (e.g. not captured
by your assay technology), or whether alternative gene symbols should be used.

### High unresolved rate

If many cells are labelled `"unknown"` after consensus, try:

```{r unresolved}
# 1. Allow even splits
consensus <- resolve_consensus_labels(
  ...,
  allow_even_split = TRUE
)

# 2. Disable ordered tie-breaking so either tie-breaker can resolve
consensus <- resolve_consensus_labels(
  ...,
  ordered_tiebreak = FALSE
)

# 3. Inspect which method combinations are causing disagreements
table(results$labels$method_1, results$labels$method_2)
```

### Collapsed broad labels

If all clusters receive the same broad label, CellVoteR retains the original
cluster structure automatically. This is expected behaviour for highly
homogeneous datasets (e.g. a sample consisting entirely of tumour cells).
In this case, the numeric cluster prefixes (e.g. `1_sc1`, `2_sc1`) trigger
testing against the full fine marker panel rather than a lineage-specific
subset.

### Large datasets

For datasets exceeding available RAM, convert the SCE to HDF5-backed storage
after `create_sce()`:

```{r hdf5}
HDF5Array::saveHDF5SummarizedExperiment(sce, dir = "my_hdf5_sce")
sce <- HDF5Array::loadHDF5SummarizedExperiment("my_hdf5_sce")
```

### Parallelisation

Key functions accept a `BPPARAM` argument for parallelisation via
[`BiocParallel`](https://bioconductor.org/packages/BiocParallel/):

```{r parallel}
library(BiocParallel)

results <- run_cellvoter(
  sce,
  annotation_args = list(
    broad_args = list(BPPARAM = MulticoreParam(4)),
    rank_args  = list(BPPARAM = MulticoreParam(4))
  )
)
```

---

## Session Info

```{r session}
sessionInfo()
```
