---
title: "Consensus and Results"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Consensus and Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This article describes how CellVoteR resolves method-level annotations into final cell labels.

# Resolve Consensus Labels

```{r resolve}
consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = FALSE,
  ordered_tiebreak  = TRUE
)
```

# Attach Labels

```{r attach}
sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method
```

# Voting Parameters

| Parameter | Default | Effect |
|---|---|---|
| `allow_even_split` | `FALSE` | When `TRUE`, a 2-of-4 plurality can be accepted |
| `ordered_tiebreak` | `TRUE` | When `TRUE`, tie-breakers are checked in priority order |
| `unassigned_label` | `"Unknown"` | Label assigned when consensus cannot be resolved |

# Inspect Results

```{r inspect}
table(sce$cellVoteR_label)
table(sce$cellVoteR_method)
table(sce$cellVoteR_label, sce$cellVoteR_method)
```

# Compare Method Outputs

```{r per-method}
table(results$labels$method_1)
table(results$labels$method_2)
table(results$labels$method_3)
table(results$labels$method_4)
table(results$labels$global_1)
table(results$labels$global_2)
```

# Method Agreement

```{r agreement}
label_df <- as.data.frame(results$labels[1:4])

mean(apply(label_df, 1, function(x) length(unique(x)) == 1L))

label_df$n_agree <- apply(label_df, 1, function(x) max(table(x)))
table(label_df$n_agree)
```

# Re-run Consensus

```{r rerun-consensus}
consensus_liberal <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = TRUE
)
```
