---
title: "Consensus and Results"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: false
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Consensus and Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```


After running the ensemble annotation methods, the next step is to resolve the results into a final set of cell labels. This is done through a configurable consensus resolution step. This step is intentionally separate from the annotation methods, allowing you to adjust voting parameters and run `resolve_consensus_labels()` without needing to re-running the entire annotation pipeline.


```{r resolve}
consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = FALSE,
  ordered_tiebreak  = TRUE
)
```

# Decision hierarchy

The process of resolving a final label for each cell is based on a simple voting scheme. Each of the four primary annotation workflows casts a "vote" for a label, and the two global tie-breakers are then used to resolve any ties.

For each cell, the following logic is applied in order:

1. **Strong majority** — if one label receives strictly more than 50% of
   method votes (3 of 4 by default) it is assigned immediately.

2. **Tie-breaker agreement** — if the two tie-breakers agree with each other
   *and* their shared label matches the leading candidate, that label is
   assigned.

3. **Ordered tie-breaking** (`ordered_tiebreak = TRUE`, default) — tie-breaker
   1 is tried first; if it matches the leading candidate the label is assigned.
   Otherwise tie-breaker 2 is tried.

4. **Either tie-breaking** (`ordered_tiebreak = FALSE`) — either tie-breaker
   agreeing with the leading candidate is sufficient, with no priority between
   them.

5. **Unresolved** — if every method disagrees (no leading candidate exists) or
   no tie-breaker resolves the split, `unassigned_label` is assigned.

# Voting parameters

| Parameter | Default | Effect |
|---|---|---|
| **allow_even_split** | **FALSE** | When **TRUE**, a 2-of-4 plurality is accepted as majority |
| **ordered_tiebreak** | **TRUE** | When **FALSE**, either tie-breaker can resolve a split |
| **unassigned_label** | **"Unknown"** | Label for unresolved cells |

# Saving labels back to the SCE

```{r attach}
sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method
```


# Inspecting Results

## Summary tables

```{r inspect}
# Final label distribution
table(sce$cellVoteR_label)

# Decision method breakdown
table(sce$cellVoteR_method)

# Cross-tabulate label vs decision method
table(sce$cellVoteR_label, sce$cellVoteR_method)
```

## Per-method label distributions

Comparing individual method outputs before consensus is useful for
understanding where methods agree and disagree:

```{r per-method}
table(results$labels$method_1)   # cluster-based, full
table(results$labels$method_2)   # cluster-based, reduced
table(results$labels$method_3)   # enrichment-based, full
table(results$labels$method_4)   # enrichment-based, reduced
table(results$labels$global_1)   # tie-breaker 1
table(results$labels$global_2)   # tie-breaker 2
```

### Method agreement

A useful diagnostic is to look at how often all four primary methods agree:

```{r agreement}
label_df <- as.data.frame(results$labels[1:4])

# Proportion of cells where all four methods agree
mean(apply(label_df, 1, function(x) length(unique(x)) == 1L))

# Per-cell method agreement count
label_df$n_agree <- apply(label_df, 1, function(x) max(table(x)))
table(label_df$n_agree)
```

## Re-running consensus with different parameters

```{r re-resolve}
# allowing even splits
consensus_liberal <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown",
  allow_even_split  = TRUE
)

table(consensus_liberal$label)

# Compare unresolved rate between settings
mean(consensus$label == "unknown")
mean(consensus_liberal$label == "unknown")
```
