---
title: "Introduction to CellVoteR"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{CellVoteR: Ensemble cell type annotation for single-cell RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```
# Overview

CellVoteR is an ensemble-based pipeline for robust cell type annotation in single-cell RNA-seq data.
Rather than assigning labels from a single marker score or clustering strategy, CellVoteR runs complementary annotation methods across two feature spaces and resolves the results through a configurable consensus step.
The workflow has four main stages:

1. Prepare a SingleCellExperiment object from raw counts.
2. Load and configure broad and fine marker sets.
3. Build full and marker-reduced analysis tracks.
4. Run ensemble annotation and resolve final consensus labels.

# Installation

```{r install}
# install.packages("devtools")
devtools::install_github("ajxa/CellVoteR")
```

```{r load}
library(CellVoteR)
```

---

# Complete Workflow

```{r complete-workflow}
markers <- load_markers(file_path = "path/to/input_markers.xlsx")

markers$broad <- build_broad_marker_config(
  marker_list       = markers$broad,
  priority_order    = c("vasculature", "immune"),
  default_threshold = 0.25
)

sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"
)

sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)
sce <- normalize_counts(sce)
sce <- prepare_sce(sce, markers)

results <- run_cellvoter(sce)

consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown"
)

sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method

table(sce$cellVoteR_label)
table(sce$cellVoteR_method)
```


<!-- ## Step 1: Preparing Marker Inputs -->

<!-- Markers are the backbone of CellVoteR's annotation strategy. They are organised into two tiers: -->

<!-- ### Broad markers (lineage-specific) -->

<!-- Broad markers define coarse cell lineages (e.g. *Immune*, *Vasculature*, -->
<!-- *Other*). Therefore, they must be: -->

<!-- - **Small** — typically 2–5 genes per category. -->

<!-- - **Mutually exclusive** — no gene should appear in more than one broad category. -->

<!-- - **Biologically diagnostic** — genes that robustly delineate lineages even in heterogeneous datasets. -->

<!-- These broad markers are loaded and then configured with `build_broad_marker_config()`, -->
<!-- which assigns expression thresholds and priority rankings used for tie-breaking -->
<!-- when a cell passes multiple broad categories. -->

<!-- ### Fine markers (cell-type specific) -->

<!-- Fine markers define sub-populations within each broad lineage (e.g. *B cell*, -->
<!-- *T cell*, *NK cell* within *Immune*). They can be larger gene sets and are -->
<!-- used for Fisher's Exact Test scoring during fine annotation. These marker do not need -->
<!-- to be mutually exclusive, but should sufficiently distinguish between to cell types from a common -->
<!-- lineage, e.g, T cells vs B cells and Mural cells vs Endothelial cells. -->


<!-- ### Loading markers -->

<!-- User-supplied markers can be loaded from either `Excel`, `CSV`, or `TXT` files. The files must be structured to comprise four columns: `type` (broad/fine), `category`, `label`, and `marker`: -->


<!-- | type | category | label | marker | -->
<!-- |---|---|---|---| -->
<!-- | broad | immune |  | PTPRC | -->
<!-- | broad | vasculature |  | CDH5 | -->
<!-- | broad | vasculature |  | VWF | -->
<!-- | fine | immune | T cell | CD2 | -->
<!-- | fine | immune | T cell | CD3D | -->
<!-- | fine | immune | T cell | IL32 | -->
<!-- | fine | immune | B cell | CD79A | -->
<!-- | fine | immune | B cell | CD79B | -->
<!-- | fine | vasculature | Mural cell | IGFBP7 | -->
<!-- | fine | vasculature | Mural cell | FN1 | -->
<!-- | fine | vasculature | Endothelial | A2M | -->
<!-- | fine | vasculature | Endothelial | IGFBP7 | -->


<!-- > When defining broad category markers, leave the **label** field blank. For fine cell type markers within that broad category, assign a **label**. -->


<!-- ```{r load-markers} -->
<!-- markers <- load_markers(file_path = "path/to/input_markers.xlsx") -->

<!-- # Inspect the structure -->
<!-- str(markers$broad)   # named list of character vectors -->
<!-- str(markers$fine)    # nested named list: broad category > fine cell type > genes -->
<!-- ``` -->

<!-- ### Configuring broad markers -->

<!-- `build_broad_marker_config()` processes the raw broad marker list, attaching -->
<!-- expression thresholds and priority ranks used during the enrichment-based -->
<!-- annotation step. -->

<!-- ```{r build-broad} -->
<!-- markers$broad <- build_broad_marker_config( -->
<!--   marker_list       = markers$broad, -->
<!--   priority_order    = c("vasculature", "immune"),  # higher priority listed first -->
<!--   default_threshold = 0.25                          # default logcounts threshold -->
<!-- ) -->

<!-- # Each broad category now has markers, expr_threshold, coexp_min, and priority -->
<!-- str(markers$broad$immune) -->
<!-- ``` -->

<!-- > The **priority_order** argument controls tie-breaking when a cell passes expression thresholds for more than one broad category - categories listed earlier receive a lower (higher priority) numeric rank. -->

<!-- --- -->

<!-- ## Step 2: Object Creation & QC -->

<!-- CellVoteR works natively with -->
<!-- [*SingleCellExperiment*](https://bioconductor.org/packages/SingleCellExperiment/) -->
<!-- objects. Use `create_sce()` to construct one from your raw data. -->

<!-- ### From (in-memory) sparse matrix -->

<!-- ```{r create-sce-memory} -->
<!-- sce <- create_sce( -->
<!--   counts        = my_sparse_matrix,  # dgCMatrix, genes x cells -->
<!--   cell_metadata = my_metadata_df     # data.frame, one row per cell (optional) -->
<!-- ) -->
<!-- ``` -->

<!-- ### From file path -->

<!-- `create_sce()` also accepts file paths, which is useful for large datasets -->
<!-- where the matrix is stored on disk: -->

<!-- ```{r create-sce-files} -->
<!-- # From RDS files -->
<!-- sce <- create_sce( -->
<!--   counts        = "path/to/counts.rds", -->
<!--   cell_metadata = "path/to/metadata.rds"   # also accepts .csv or .tsv -->
<!-- ) -->

<!-- # From MTX triplet files -->
<!-- sce <- create_sce( -->
<!--   mtx_file   = "path/to/matrix.mtx.gz", -->
<!--   cells_file = "path/to/barcodes.tsv", -->
<!--   genes_file = "path/to/features.tsv" -->
<!-- ) -->
<!-- ``` -->

<!-- --- -->

<!-- ## Step 3: Quality Control -->

<!-- `assess_cell_quality()` calculates per-cell QC metrics and optionally removes -->
<!-- low-quality cells before downstream analysis. -->

<!-- ```{r qc} -->
<!-- sce <- assess_cell_quality(sce, remove_failed_cells = TRUE) -->
<!-- ``` -->

<!-- > Cells that fail QC are flagged in **colData(sce)$QC_PASS**. Setting **remove_failed_cells = TRUE** subsets the object to passing cells only. -->

<!-- --- -->

<!-- ## Step 4: Normalisation -->

<!-- CellVoteR uses a pooling-based normalisation strategy -->
<!-- ([Lun et al. 2016](https://doi.org/10.1186/s13059-016-0947-7)) via -->
<!-- `scran::computePooledFactors`, followed by log-normalisation with -->
<!-- `scuttle::logNormCounts`. -->

<!-- ```{r normalise} -->
<!-- sce <- normalize_counts(sce) -->
<!-- ``` -->

<!-- > This adds a *logcounts* assay to the SCE and sets `sizeFactors()`. The *logcounts* assay is required by all downstream steps. -->

<!-- --- -->

<!-- ## Step 5: Building Analysis Tracks -->

<!-- The `prepare_sce()` function co-ordinates the key preprocessing step. This function performs/applies, -->
<!-- the following calculations and logic: -->

<!-- 1. Validates broad and fine marker configurations against the expression matrix. -->

<!-- 2. Builds two independent feature spaces — the **full HVG space** and the **reduced marker-defined space**. -->

<!-- 3. Runs PCA and unsupervised clustering (Leiden via SNN graph) on each space. -->

<!-- 4. Attaches the marker configuration and filtered fine markers to the SCE metadata. -->

<!-- 5. Stores the reduced feature space as an **altExp** named *"user_panel"*. -->

<!-- ```{r prepare} -->
<!-- sce <- prepare_sce(sce, markers) -->
<!-- ``` -->

<!-- After this step, the initial SingleCellExepriment object structure is as follows: -->

<!-- ``` -->
<!-- sce -->
<!-- ├── assays: counts, logcounts -->
<!-- ├── rowSubset("broad_hvgs")           ←    HVGs used for broad clustering -->
<!-- ├── reducedDim("PCA_broad_hvg")       ←    PCA on full HVG space -->
<!-- ├── colData$cluster_broad_hvg         ←    Leiden clusters (full space) -->
<!-- ├── colData$cluster_broad_reduced     ←    Leiden clusters (reduced space) -->
<!-- ├── metadata$marker_config            ←    full marker configuration -->
<!-- ├── metadata$filtered_fine_markers    ←    fine markers present in data -->
<!-- ├── metadata$missing_by_label         ←    per-label missing marker report -->
<!-- └── altExp("user_panel")              ←    reduced feature SCE -->
<!--     ├── assays: counts, logcounts -->
<!--     ├── reducedDim("PCA") -->
<!--     ├── colData$cluster -->
<!--     ├── metadata$marker_config -->
<!--     ├── metadata$filtered_fine -->
<!--     └── metadata$params -->
<!-- ``` -->

<!-- ### Automatic parameter estimation -->

<!-- Clustering parameters (number of PCs, SNN neighbourhood size $k$, Leiden -->
<!-- resolution) are estimated automatically from cell count using bounded -->
<!-- $log/sqrt$ scaling via `estimate_cluster_params()`. However, these parameters can be overridden if required: -->

<!-- ```{r prepare-override} -->
<!-- sce <- prepare_sce( -->
<!--   sce, -->
<!--   markers, -->
<!--   n_hvgs     = 3000L, -->
<!--   n_pcs      = 30L, -->
<!--   k          = 20L, -->
<!--   resolution = 0.8 -->
<!-- ) -->
<!-- ``` -->

<!-- ### Marker overlap reporting -->

<!-- In some cases, you may have fine markers that are partially missing from your dataset: the `prepare_sce()` function captures this information and details the labels that are most affected. which can be inspected as follows: -->

<!-- ```{r missing-markers} -->
<!-- # Inspect missing marker report after prepare_sce() -->
<!-- metadata(sce)$missing_by_label -->
<!-- ``` -->

<!-- Missing (in the data) marker are automatically removed from the fine marker sets used for scoring, so annotation still proceeds with the genes that are present. This only occurs when the total number of missing markers is <50% of the total distinct markers supplied - this parameter can be relaxed or tightened accordingly, by altering the `overlap_feat_percent` argument: -->

<!-- ```{r override-marker-overlap} -->

<!-- prepare_sce( -->
<!--   sce, -->
<!--   markers, -->
<!--   overlap_feat_percent = 75       # more stringent -->
<!-- ) -->
<!-- ``` -->


<!-- --- -->

<!-- ## Step 6: Annotation Methods -->

<!-- The `run_cellvoter()` orchestrates all of the individual annotation pipelines and returns a -->
<!-- named list of per-cell label factors, one per method: -->

<!-- ```{r run} -->
<!-- results <- run_cellvoter(sce) -->
<!-- ``` -->

# What Happens Internally?

At a high level, CellVoteR compares four primary annotation methods and two global tie-breakers.
The four primary methods combine:

- two feature spaces: full highly variable gene space and reduced marker-panel space
- two broad annotation strategies: cluster-based and enrichment-based

The consensus step then assigns a final label when the methods agree strongly, or uses the global tie-breakers to resolve partial disagreement.

# Where to Go Next?

For details on input file formats and marker structure, see vignette("user_inputs").
For preprocessing, analysis tracks, and prepare_sce(), see the preprocessing article.
For the annotation methods and custom parameters, see the annotation methods article.
For consensus voting, output inspection, and troubleshooting, see the consensus and results articles.





<!-- ### What this does internally -->

<!-- Four primary methods and two global tie-breakers are run: -->

<!-- | Method | Feature space | Broad strategy | Subcluster feature mode | -->
<!-- |---|---|---|---| -->
<!-- | Method 1 | Full (HVG) | Cluster-based | HVG | -->
<!-- | Method 2 | Reduced (panel) | Cluster-based | All | -->
<!-- | Method 3 | Full (HVG) | Enrichment-based | HVG | -->
<!-- | Method 4 | Reduced (panel) | Enrichment-based | All | -->
<!-- | Tie-breaker 1 | Full (HVG) | None (global) | — | -->
<!-- | Tie-breaker 2 | Reduced (panel) | None (global) | — | -->

<!-- Each primary method follows the same pipeline: -->

<!-- ``` -->
<!-- broad annotation -->
<!--       ↓ -->
<!-- subcluster_labels() -->
<!--       ↓ -->
<!-- rank_cluster_markers()          ←   DE testing per sub-cluster -->
<!--       ↓ -->
<!-- extract_top_markers()           ←   select top N genes per cluster -->
<!--       ↓ -->
<!-- score_markers_against_panel()   ←   Fisher's Exact Test + overlap similarity -->
<!--       ↓ -->
<!-- assign_fine_labels()            ←   best label per cluster    →    mapped to cells -->
<!-- ``` -->

<!-- ### Broad annotation strategies -->

<!-- **Cluster-based** (`annotate_broad_clusters`): Runs DE testing on pre-existing -->
<!-- unsupervised clusters. Each cluster is assigned the broad category whose -->
<!-- curated markers have the lowest median rank among significantly up-regulated -->
<!-- genes (FDR ≤ 0.05, AUC ≥ 0.6 by default). -->

<!-- **Enrichment-based** (`annotate_broad_cells`): Assigns labels directly to -->
<!-- individual cells by aggregating expression across each broad category's marker -->
<!-- genes and comparing against category-specific thresholds. Does not depend on -->
<!-- clustering. -->

<!-- > In some datasets (highly homogeneous tumour sample), all clusters may -->
<!-- receive the same broad label. CellVoteR detects this and retains the original -->
<!-- cluster structure for subclustering rather than collapsing to a single group, -->
<!-- preserving the fine-resolution information for downstream processes. -->

<!-- ### Annotation Method Results -->

<!-- After running `run_cellvoter()`, the returned SingleCellExeriment object is populated with all of th intermediate cluster labels, which can be accessed by querying the colData() columns of the object: -->

<!-- | Column | Description | -->
<!-- |---|---| -->
<!-- | **cluster_broad_hvg** | Pre-existing HVG clusters from `prepare_sce()` | -->
<!-- | **cluster_broad_reduced** | Pre-existing reduced clusters from `prepare_sce()` | -->
<!-- | **broad_cluster_m1** | Broad labels, method 1 | -->
<!-- | **broad_cluster_sub_m1** | Sub-cluster labels, method 1 | -->
<!-- | **broad_cluster_m2** | Broad labels, method 2 | -->
<!-- | **broad_cluster_sub_m2** | Sub-cluster labels, method 2 | -->
<!-- | **broad_enrichment_m3** | Broad labels, method 3 | -->
<!-- | **broad_enrichment_sub_m3** | Sub-cluster labels, method 3 | -->
<!-- | **broad_enrichment_m4** | Broad labels, method 4 | -->
<!-- | **broad_enrichment_sub_m4** | Sub-cluster labels, method 4 | -->

<!-- ### Customising parameters -->

<!-- The main `run_cellvoter()` function accepts an `annotation_args` list which can be used to customise various parameters of the underlying internal functions, if required. There following parameter lists that can be specified, which each control specific aspects of the underlying logic: -->


<!-- - **rank_args** Controls parameters associated with `rank_cluster_markers()`:  -->
<!--   - assay_type -->
<!--   - test_type -->
<!--   - direction -->
<!--   - pval_type -->
<!--   - min_prop -->
<!--   - BPPARAM -->

<!-- - **broad_args** is identical to the `rank_args`, but only controls the ranking inside the  -->
<!-- `annotate_broad_clusters` function. This is useful for altering the parameters which control how the  -->
<!-- broad cell lineage labels (e.g, immune, vasculature) are defined - these broad markers are small and highly specific and so you may wish to use a more lenient `min_prop` compared to the fine labels and this allows for such fine control of the process.  -->

<!-- - **extract_args** Controls parameters associated with `extract_top_markers()`:  -->
<!--   - fdr_threshold -->
<!--   - effect_threshold -->
<!--   - target_n -->

<!-- An following is an example of how this can be used to independently alter the broad and fine labelling logic: -->

<!-- ```{r run-custom} -->
<!-- results <- run_cellvoter( -->
<!--   sce, -->
<!--   return_full_output = TRUE, -->
<!--   annotation_args = list( -->

<!--     # Controls ranking inside annotate_broad_clusters (methods 1 and 2) -->
<!--     # Lenient min_prop appropriate for small, specific broad marker sets -->
<!--     broad_args = list( -->
<!--       test_type = "wilcox", -->
<!--       min_prop  = 0.1 -->
<!--     ), -->

<!--     # Controls ranking inside .run_fine_annotation (all six methods) -->
<!--     # Stricter settings for fine sub-cluster marker extraction -->
<!--     rank_args = list( -->
<!--       test_type = "wilcox", -->
<!--       min_prop  = 0.25 -->
<!--     ), -->

<!--     # Controls top marker extraction -->
<!--     extract_args = list( -->
<!--       fdr_threshold    = 0.05, -->
<!--       effect_threshold = 0.6, -->
<!--       target_n         = 100L -->
<!--     ) -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->

<!-- > **broad_args** vs **rank_args**: these both control `rank_cluster_markers()` but at different pipeline stages. *broad_args* affects how broad lineage labels are assigned (methods 1 and 2 only); *rank_args* affects how sub-cluster marker genes are extracted for Fisher scoring (all six methods). Methods 3 and 4 use *rank_args* only, as the enrichment-based broad step does not call *rank_cluster_markers()*. -->

<!-- ### Accessing full outputs -->

<!-- > Setting `return_full_output` = **TRUE**, returns the per-cluster Fisher scores and similarity -->
<!-- values are for every method (by default this is set to *FALSE*). -->

<!-- ```{r full-output} -->
<!-- # Per-cluster score table for method 1 -->
<!-- results$full_output$method_1$scores -->

<!-- # Per-cluster score table for tie-breaker 2 -->
<!-- results$full_output$global_2$scores -->
<!-- ``` -->

<!-- --- -->

<!-- ## Step 7: Resolving Consensus Labels -->

<!-- Consensus resolution is intentionally a separate step. This means you can -->
<!-- adjust voting parameters and re-run `resolve_consensus_labels()` as many -->
<!-- times as needed without re-running the annotation pipeline. -->

<!-- ```{r resolve} -->
<!-- consensus <- resolve_consensus_labels( -->
<!--   label_list        = results$labels, -->
<!--   method_names      = results$method_names, -->
<!--   tie_breaker_names = results$tie_breaker_names, -->
<!--   unassigned_label  = "unknown", -->
<!--   allow_even_split  = FALSE, -->
<!--   ordered_tiebreak  = TRUE -->
<!-- ) -->
<!-- ``` -->

<!-- ### Decision hierarchy -->

<!-- For each cell, the following logic is applied in order: -->

<!-- 1. **Strong majority** — if one label receives strictly more than 50% of -->
<!--    method votes (3 of 4 by default) it is assigned immediately. -->

<!-- 2. **Tie-breaker agreement** — if the two tie-breakers agree with each other -->
<!--    *and* their shared label matches the leading candidate, that label is -->
<!--    assigned. -->

<!-- 3. **Ordered tie-breaking** (`ordered_tiebreak = TRUE`, default) — tie-breaker -->
<!--    1 is tried first; if it matches the leading candidate the label is assigned. -->
<!--    Otherwise tie-breaker 2 is tried. -->

<!-- 4. **Either tie-breaking** (`ordered_tiebreak = FALSE`) — either tie-breaker -->
<!--    agreeing with the leading candidate is sufficient, with no priority between -->
<!--    them. -->

<!-- 5. **Unresolved** — if every method disagrees (no leading candidate exists) or -->
<!--    no tie-breaker resolves the split, `unassigned_label` is assigned. -->

<!-- ### Voting parameters -->

<!-- | Parameter | Default | Effect | -->
<!-- |---|---|---| -->
<!-- | **allow_even_split** | **FALSE** | When **TRUE**, a 2-of-4 plurality is accepted as majority | -->
<!-- | **ordered_tiebreak** | **TRUE** | When **FALSE**, either tie-breaker can resolve a split | -->
<!-- | **unassigned_label** | **"Unknown"** | Label for unresolved cells | -->

<!-- ### Attaching labels to the SCE -->

<!-- ```{r attach} -->
<!-- sce$cellVoteR_label  <- consensus$label -->
<!-- sce$cellVoteR_method <- consensus$method -->
<!-- ``` -->

<!-- --- -->

<!-- ## Step 8: Inspecting Results -->

<!-- ### Summary tables -->

<!-- ```{r inspect} -->
<!-- # Final label distribution -->
<!-- table(sce$cellVoteR_label) -->

<!-- # Decision method breakdown -->
<!-- table(sce$cellVoteR_method) -->

<!-- # Cross-tabulate label vs decision method -->
<!-- table(sce$cellVoteR_label, sce$cellVoteR_method) -->
<!-- ``` -->

<!-- ### Per-method label distributions -->

<!-- Comparing individual method outputs before consensus is useful for -->
<!-- understanding where methods agree and disagree: -->

<!-- ```{r per-method} -->
<!-- table(results$labels$method_1)   # cluster-based, full -->
<!-- table(results$labels$method_2)   # cluster-based, reduced -->
<!-- table(results$labels$method_3)   # enrichment-based, full -->
<!-- table(results$labels$method_4)   # enrichment-based, reduced -->
<!-- table(results$labels$global_1)   # tie-breaker 1 -->
<!-- table(results$labels$global_2)   # tie-breaker 2 -->
<!-- ``` -->

<!-- ### Method agreement -->

<!-- A useful diagnostic is to look at how often all four primary methods agree: -->

<!-- ```{r agreement} -->
<!-- label_df <- as.data.frame(results$labels[1:4]) -->

<!-- # Proportion of cells where all four methods agree -->
<!-- mean(apply(label_df, 1, function(x) length(unique(x)) == 1L)) -->

<!-- # Per-cell method agreement count -->
<!-- label_df$n_agree <- apply(label_df, 1, function(x) max(table(x))) -->
<!-- table(label_df$n_agree) -->
<!-- ``` -->

<!-- ### Re-running consensus with different parameters -->

<!-- ```{r re-resolve} -->
<!-- # allowing even splits -->
<!-- consensus_liberal <- resolve_consensus_labels( -->
<!--   label_list        = results$labels, -->
<!--   method_names      = results$method_names, -->
<!--   tie_breaker_names = results$tie_breaker_names, -->
<!--   unassigned_label  = "unknown", -->
<!--   allow_even_split  = TRUE -->
<!-- ) -->

<!-- table(consensus_liberal$label) -->

<!-- # Compare unresolved rate between settings -->
<!-- mean(consensus$label == "unknown") -->
<!-- mean(consensus_liberal$label == "unknown") -->
<!-- ``` -->

<!-- --- -->

<!-- # Complete Workflow -->

<!-- The full pipeline from raw data to annotated SCE: -->

<!-- ```{r full-workflow} -->
<!-- library(CellVoteR) -->

<!-- # ── 1. Load and configure markers ────────────────────────────────────────────── -->
<!-- markers <- load_markers(file_path = "path/to/input_markers.xlsx") -->

<!-- markers$broad <- build_broad_marker_config( -->
<!--   marker_list       = markers$broad, -->
<!--   priority_order    = c("vasculature", "immune"), -->
<!--   default_threshold = 0.25 -->
<!-- ) -->

<!-- # ── 2. Create SCE ────────────────────────────────────────────────────────────── -->
<!-- sce <- create_sce( -->
<!--   counts        = "path/to/counts.rds", -->
<!--   cell_metadata = "path/to/metadata.rds" -->
<!-- ) -->

<!-- # ── 3. QC ────────────────────────────────────────────────────────────────────── -->
<!-- sce <- assess_cell_quality(sce, remove_failed_cells = TRUE) -->

<!-- # ── 4. Normalise ─────────────────────────────────────────────────────────────── -->
<!-- sce <- normalize_counts(sce) -->

<!-- # ── 5. Build analysis tracks ─────────────────────────────────────────────────── -->
<!-- sce <- prepare_sce(sce, markers) -->

<!-- # ── 6. Run ensemble annotation ───────────────────────────────────────────────── -->
<!-- results <- run_cellvoter(sce) -->

<!-- # ── 7. Resolve consensus ─────────────────────────────────────────────────────── -->
<!-- consensus <- resolve_consensus_labels( -->
<!--   label_list        = results$labels, -->
<!--   method_names      = results$method_names, -->
<!--   tie_breaker_names = results$tie_breaker_names, -->
<!--   unassigned_label  = "unknown" -->
<!-- ) -->

<!-- # ── 8. Attach labels ─────────────────────────────────────────────────────────── -->
<!-- sce$cellVoteR_label  <- consensus$label -->
<!-- sce$cellVoteR_method <- consensus$method -->

<!-- # ── 9. Inspect ───────────────────────────────────────────────────────────────── -->
<!-- table(sce$cellVoteR_label) -->
<!-- table(sce$cellVoteR_method) -->
<!-- ``` -->

<!-- --- -->

<!-- # Tips and Troubleshooting -->

<!-- ## Low marker overlap -->

<!-- If `prepare_sce()` warns about low fine marker overlap, inspect which labels -->
<!-- are most affected: -->

<!-- ```{r missing} -->
<!-- metadata(sce)$missing_by_label -->
<!-- ``` -->

<!-- Consider whether the missing genes are platform-specific (e.g. not captured -->
<!-- by your assay technology), or whether alternative gene symbols should be used. -->

<!-- ### High unresolved rate -->

<!-- If many cells are labelled `"unknown"` after consensus, try: -->

<!-- ```{r unresolved} -->
<!-- # 1. Allow even splits -->
<!-- consensus <- resolve_consensus_labels( -->
<!--   ..., -->
<!--   allow_even_split = TRUE -->
<!-- ) -->

<!-- # 2. Disable ordered tie-breaking so either tie-breaker can resolve -->
<!-- consensus <- resolve_consensus_labels( -->
<!--   ..., -->
<!--   ordered_tiebreak = FALSE -->
<!-- ) -->

<!-- # 3. Inspect which method combinations are causing disagreements -->
<!-- table(results$labels$method_1, results$labels$method_2) -->
<!-- ``` -->

<!-- ## Collapsed broad labels -->

<!-- If all clusters receive the same broad label, CellVoteR retains the original -->
<!-- cluster structure automatically. This is expected behaviour for highly -->
<!-- homogeneous datasets (e.g. a sample consisting entirely of tumour cells). -->
<!-- In this case, the numeric cluster prefixes (e.g. `1_sc1`, `2_sc1`) trigger -->
<!-- testing against the full fine marker panel rather than a lineage-specific -->
<!-- subset. -->

<!-- ## Large datasets -->

<!-- For datasets exceeding available RAM, convert the SCE to HDF5-backed storage -->
<!-- after `create_sce()`: -->

<!-- ```{r hdf5} -->
<!-- HDF5Array::saveHDF5SummarizedExperiment(sce, dir = "my_hdf5_sce") -->
<!-- sce <- HDF5Array::loadHDF5SummarizedExperiment("my_hdf5_sce") -->
<!-- ``` -->

<!-- ## Parallelisation -->

<!-- Key functions accept a `BPPARAM` argument for parallelisation via -->
<!-- [`BiocParallel`](https://bioconductor.org/packages/BiocParallel/): -->

<!-- ```{r parallel} -->
<!-- library(BiocParallel) -->

<!-- results <- run_cellvoter( -->
<!--   sce, -->
<!--   annotation_args = list( -->
<!--     broad_args = list(BPPARAM = MulticoreParam(4)), -->
<!--     rank_args  = list(BPPARAM = MulticoreParam(4)) -->
<!--   ) -->
<!-- ) -->
<!-- ``` -->

<!-- --- -->

# Session Info

```{r session, echo=FALSE, eval=TRUE}
print(sessionInfo())
```
