---
title: "Introduction to CellVoteR"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
  self_contained: false
  default_style: "light"
  downcute_theme: "default"
  code_folding: show
  toc_depth: 4
  toc_float: true
vignette: >
  %\VignetteIndexEntry{CellVoteR: Ensemble cell type annotation for single-cell RNA-seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```
# Overview

CellVoteR is an ensemble-based pipeline for robust cell type annotation in single-cell RNA-seq data.

Rather than assigning labels from a single marker score or clustering strategy, CellVoteR runs complementary annotation methods across two feature spaces and resolves the results through a configurable consensus step.
The workflow has four main stages:

1. Prepare a SingleCellExperiment object from raw counts.
2. Load and configure broad and fine marker sets.
3. Build full and marker-reduced analysis tracks.
4. Run ensemble annotation and resolve final consensus labels.

# Installation

```{r install}
# install.packages("devtools")
devtools::install_github("ajxa/CellVoteR")
```

```{r load}
library(CellVoteR)
```

---

# Complete Workflow

```{r complete-workflow}
markers <- load_markers(file_path = "path/to/input_markers.xlsx")

markers$broad <- build_broad_marker_config(
  marker_list       = markers$broad,
  priority_order    = c("vasculature", "immune"),
  default_threshold = 0.25
)

sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"
)

sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)
sce <- normalize_counts(sce)
sce <- prepare_sce(sce, markers)

results <- run_cellvoter(sce)

consensus <- resolve_consensus_labels(
  label_list        = results$labels,
  method_names      = results$method_names,
  tie_breaker_names = results$tie_breaker_names,
  unassigned_label  = "unknown"
)

sce$cellVoteR_label  <- consensus$label
sce$cellVoteR_method <- consensus$method

table(sce$cellVoteR_label)
table(sce$cellVoteR_method)
```




# What Happens Internally?

At a high level, CellVoteR compares four primary annotation methods and two global tie-breakers.
The four primary methods combine:

- two feature spaces: full highly variable gene space and reduced marker-panel space
- two broad annotation strategies: cluster-based and enrichment-based

The consensus step then assigns a final label when the methods agree strongly, or uses the global tie-breakers to resolve partial disagreement.

# Where to Go Next?

- For details on input file formats and marker structure, see [User inputs and required formats](user_inputs.html).

- For information on preprocessing, normalisation and preparing the analysis tracks, see [Preprocessing and analysis tracks](preprocessing.html).

- For information on the individual annotation methods, see [Annotation methods](annotation_methods.html) .

- For details on the consensus resolution step, see [Consensus and results](consensus_and_results.html).

- For any additional practical guidance, including troubleshooting tips and common issues, see [Troubleshooting](troubleshooting.html).
