---
title: "Preprocessing and Analysis Tracks"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Preprocessing and Analysis Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This article describes the preprocessing steps used before running CellVoteR annotation.

The main steps are:

1. Assess cell quality.
2. Normalise counts.
3. Build full and reduced analysis tracks with `prepare_sce()`.

# Create a SingleCellExperiment

This is described in more detail in the [User inputs and required formats](user_inputs.html) article and is done by running `create_sce()` which can accept multiple data formats.

```{r create-sce}
sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"
)
```

# Quality Control

After creating the `SingleCellExperiment`, the `assess_cell_quality()` helper function can be used to calculate per-cell QC metrics and optionally remove low-quality cells before downstream analysis. This is usually the first step after creating the SCE, as it ensures that downstream processes are not affected by low-quality cells that may have aberrant expression profiles. 

```{r qc}
sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)
```

> If the **remove_failed_cells** argument is set to **FALSE**, the QC metrics are still calculated and attached to the SCE, but no cells are removed. This allows you to inspect the QC metrics and decide on thresholds for filtering before actually removing any cells.

# Data Normalisation

Following quality control, the raw counts matrix must be normalised to account for differences in sequencing depth and other technical factors. This is a critical step, as the annotation methods rely on accurate expression values to score marker gene enrichment. CellVoteR uses a pooling-based normalisation strategy
([Lun et al. 2016](https://doi.org/10.1186/s13059-016-0947-7)) via `scran::computePooledFactors`, followed by log-normalisation with `scuttle::logNormCounts`. This is implemented in the `normalize_counts()` helper function, which takes the raw SCE and returns a new SCE with normalised logcounts.

```{r normalise}
sce <- normalize_counts(sce)
```


> The function adds a *logcounts* assay to the input SCE which contains the normalised expression values. The original raw counts are retained in the *counts* assay for reference and potential use in downstream processes that require raw counts (e.g. DE testing). The normalised logcounts are used for all downstream analyses, including HVG selection, PCA, clustering, and marker scoring.


# Build Analysis Tracks

The final step before annotation is to build the analysis tracks - the feature spaces and cluster structures on which the annotation methods depend. This is done by the `prepare_sce()` function, which takes the normalised SCE and the marker configuration as inputs and returns an SCE with all necessary metadata, reduced dimensions, and cluster assignments attached for downstream annotation.

Internally, the `prepare_sce()` function applies a series of processing steps:

1. Validates broad and fine marker configurations against the expression matrix.

2. Builds two independent feature spaces — the **full HVG space** and the **reduced marker-defined space**.
   
3. Runs PCA and unsupervised clustering (Leiden via SNN graph) on each space.

4. Attaches the marker configuration and filtered fine markers to the SCE metadata.

5. Stores the reduced feature space as an **altExp** named *"user_panel"*.

```{r prepare}
sce <- prepare_sce(sce, markers)
```

The resulting SCE contains two parallel cluster structures: one based on the full HVG space, and one based on the reduced marker-defined space. Both are used for different annotation methods, as some methods rely on broad clusters defined in the full HVG space, while others use clusters defined in the reduced space to ensure they are driven by the supplied markers.

The reduced marker-defined space is stored within the same SCE object as an alternative experiment (altExp) named *"user_panel"*. This altExp contains the log-normalised expression values for the subset of genes that overlap with the supplied markers, and is used for PCA and clustering in the reduced feature space.

The overall structure of the SCE after running `prepare_sce()` is as follows:

```
sce
├── assays: counts, logcounts
├── rowSubset("broad_hvgs")           ←    HVGs used for broad clustering
├── reducedDim("PCA_broad_hvg")       ←    PCA on full HVG space
├── colData$cluster_broad_hvg         ←    Leiden clusters (full space)
├── colData$cluster_broad_reduced     ←    Leiden clusters (reduced space)
├── metadata$marker_config            ←    full marker configuration
├── metadata$filtered_fine_markers    ←    fine markers present in data
├── metadata$missing_by_label         ←    per-label missing marker report
└── altExp("user_panel")              ←    reduced feature SCE
    ├── assays: counts, logcounts
    ├── reducedDim("PCA")
    ├── colData$cluster
    ├── metadata$marker_config
    ├── metadata$filtered_fine
    └── metadata$params
```

# Automatic Parameter Estimation

During the this final preparation step, key parameters for PCA and clustering are estimated automatically based on the number of cells in the dataset, using bounded $log/sqrt$ scaling via the `estimate_cluster_params()` helper. This ensures that the feature spaces and cluster structures are appropriately sized for the dataset at hand, without requiring manual tuning of parameters. However, if you wish to override the automatically estimated parameters, you can do so by passing custom values for the number of HVGs, PCs, SNN neighbourhood size $k$, and Leiden resolution directly to `prepare_sce()` as follows:


```{r prepare-custom}
sce <- prepare_sce(
  sce,
  markers,
  n_hvgs     = 3000L,
  n_pcs      = 30L,
  k          = 20L,
  resolution = 0.8
)
```

# Marker Overlap Reporting

After preprocessing, the `SingleCellExperiment` contains the assays, reduced dimensions, cluster labels, marker configuration, and reduced marker-panel experiment needed by the annotation workflow.

In order to ensure that the annotation methods have sufficient information to work with, `prepare_sce()` checks the supplied marker lists against the expression matrix and reports on the degree of overlap. If a large proportion of fine markers are missing from the dataset, this may indicate an issue with gene naming conventions or platform-specific gene capture, which could affect annotation performance. The function captures this information and details the labels that are most affected, which can be inspected as follows:


```{r missing-markers}
metadata(sce)$missing_by_label
```

Markers that are missing from the dataset are automatically removed from the fine marker sets used for scoring, so annotation still proceeds with the genes that are present. This only occurs when the total number of missing markers is <50% (default )of the total distinct markers supplied - this parameter can be relaxed or tightened accordingly, by altering the `overlap_feat_percent` argument:

```{r override-marker-overlap}

prepare_sce(
  sce,
  markers,
  overlap_feat_percent = 75       # more stringent
)
```

# Next steps

At this point, the `SingleCellExperiment` is fully prepared for annotation. The next step is to run the ensemble annotation methods using `run_cellvoter()`, which will generate per-method and consensus labels for each cell based on the supplied markers and cluster structures. For more details on the annotation methods, see the [Annotation methods](annotation_methods.html) article.
