---
title: "Preprocessing and Analysis Tracks"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Preprocessing and Analysis Tracks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This article describes the preprocessing steps used before running CellVoteR annotation.

The main steps are:

1. Create a `SingleCellExperiment` object.
2. Assess cell quality.
3. Normalise counts.
4. Build full and reduced analysis tracks with `prepare_sce()`.

# Create a SingleCellExperiment

```{r create-sce}
sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"
)
```

# Quality Control

```{r qc}
sce <- assess_cell_quality(sce, remove_failed_cells = TRUE)
```

# Normalisation

```{r normalise}
sce <- normalize_counts(sce)
```

# Build Analysis Tracks

```{r prepare}
sce <- prepare_sce(sce, markers)
```

# Automatic Parameter Estimation

`prepare_sce()` can estimate clustering and dimensionality-reduction parameters automatically based on dataset size.

```{r prepare-custom}
sce <- prepare_sce(
  sce,
  markers,
  n_hvgs     = 3000L,
  n_pcs      = 30L,
  k          = 20L,
  resolution = 0.8
)
```

# Marker Overlap Reporting

```{r missing-markers}
metadata(sce)$missing_by_label
```

# Resulting Object Structure

After preprocessing, the `SingleCellExperiment` contains the assays, reduced dimensions, cluster labels, marker configuration, and reduced marker-panel experiment needed by the annotation workflow.
