---
title: "User Inputs and Required Formats"
author: "Shoaib Ajaib"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "light"
    downcute_theme: "default"
    code_folding: show
    toc_depth: 4
    toc_float: true
vignette: >
  %\VignetteIndexEntry{User Inputs and Required Formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

<br>

> In order to run the complete CellVoteR workflow, you must supply two inputs:
>
> 1. Raw gene-by-cell counts matrix (sparse `dgCMatrix`, `RDS`, or `MTX triplet` file).
>   
> 2. Marker configuration — a structured list of broad and fine cell type marker genes.


# Preparing input markers

Marker panels are the backbone of CellVoteR's annotation strategy. These are organised into two tiers:

## Broad markers (lineage-specific)

Broad markers are used to define coarse cell lineages (e.g. *Immune*, *Vasculature*, *Other*) and therefore, must be:

- **Small** — typically 2–5 genes per category.

- **Mutually exclusive** — no gene should appear in more than one broad category.

- **Biologically diagnostic** — genes that robustly delineate lineages even in heterogeneous datasets.

These broad markers are loaded and then configured with `build_broad_marker_config()`, which assigns expression thresholds and priority rankings used for tie-breaking when a cell passes multiple broad categories.

## Fine markers (cell-type specific)

Fine markers define sub-populations that correspond directly to each broad lineage (e.g. *B cell*,
*T cell*, *NK cell* within the *Immune* broad marker category). They marker sets can be larger and more redundant than broad markers, as they are only used for Fisher's Exact Test scoring during fine annotation. Moreover, they do not need to be mutually exclusive, but should still sufficiently distinguish between to distinct cell types that come from the same broad lineage, e.g, T cells vs. B cells and Mural cells vs Endothelial cells.


## Loading markers

User-supplied markers can be loaded from either `Excel`, `CSV`, `TXT`, `RDS` files or supplied 
directly in the form of an R `data.frame`. Irrespective of format, the file must be structured to contain both broad and fine markers, with the following four columns:

>
- **type** — indicates whether the marker is a *broad* or *fine* 
>
- **category** — for broad markers, this is the coarse lineage category; for fine markers, this is the same as the broad category they belong to.
>
- **label** - for broad markers, this field is left blank; for fine markers, this is the specific cell type label that the marker corresponds to.
>
- **marker** — the gene symbol of the marker gene.

<br>

An example of the required format is as follows:

| type | category | label | marker |
|---|---|---|---|
| broad | immune |  | PTPRC |
| broad | vasculature |  | CDH5 |
| broad | vasculature |  | VWF |
| fine | immune | T cell | CD2 |
| fine | immune | T cell | CD3D |
| fine | immune | T cell | IL32 |
| fine | immune | B cell | CD79A |
| fine | immune | B cell | CD79B |
| fine | vasculature | Mural cell | IGFBP7 |
| fine | vasculature | Mural cell | FN1 |
| fine | vasculature | Endothelial | A2M |
| fine | vasculature | Endothelial | IGFBP7 |


> When defining broad category markers, there is no need to assign a **label** as these are only used for coarse lineage assignment. However, for the fine markers, the **label** field is required as these are used to assign the final cell type labels to clusters and cells.


```{r load-markers}
markers <- load_markers(file_path = "path/to/input_markers.xlsx")

# Inspect the structure
str(markers$broad)   # named list of character vectors
str(markers$fine)    # nested named list: broad category > fine cell type > genes
```


## Internal marker panels
CellVoteR also ships with a set of internal marker panels, which can be used as-is or modified to suit your dataset. These are stored in the `marker_panels` dataset and can be accessed as follows:

```{r internal-markers}
names(marker_panels)
names(marker_panels$GBM)

```

Currently, the only internal marker panels provided relate to glioblastoma (GBM) datasets, but additional panels may be added in future updates. However, these panels can serve as a starting point for users to create their own custom marker panels for other tissue types and disease contexts.

The internal marker panels are structured in the same way as user-supplied markers, with broad and fine markers organised into a nested list and can be directly supplied to the `load_markers()` function or explicitly by setting the `markers` parameter to the required internal marker list:

```{r internal-markers-load}
markers <- load_markers(marker_panels$GBM$gbmap_neftel_full)
markers <- load_markers(markers = marker_panels$GBM$gbmap_neftel_full)
```

## Configuring broad markers

The `build_broad_marker_config()` function processes the raw broad marker list, attaching expression thresholds and priority ranks used during the enrichment-based annotation step. 

```{r build-broad}

markers$broad <- build_broad_marker_config(
  marker_list       = markers$broad,
  priority_order    = c("vasculature", "immune"),   # higher priority listed first
  default_threshold = 0.25                          # default logcounts threshold
)

str(markers$broad$immune)

```

> The **priority_order** argument is used to assign numeric priority ranks to each broad category, which is used for tie-breaking when a cell passes expression thresholds for more than one broad category. Categories listed earlier receive a lower (higher priority) numeric rank.

# Preparing expression data

CellVoteR is designed to work natively with [*SingleCellExperiment*](https://bioconductor.org/packages/SingleCellExperiment/) (SCE)
objects. The `create_sce()` helper function can be used to construct a basic SCE from a variety of input formats, which is then processed and annotated by the downstream functions.

## From sparse matrix (in-memory)

The most direct way to create an SCE is from an in-memory sparse matrix of class `dgCMatrix` (genes x cells), with optional cell metadata as a `data.frame`:

```{r create-sce-memory}
sce <- create_sce(
  counts        = my_sparse_matrix,  # dgCMatrix, genes x cells
  cell_metadata = my_metadata_df     # data.frame, one row per cell (optional)
)
```

## From a file path

This function also accepts file paths, which is useful for large datasets where the matrix is stored on disk: currently, disk-backed matrices are not supported but this is a planned feature for a future release:

```{r create-sce-files}

sce <- create_sce(
  counts        = "path/to/counts.rds",
  cell_metadata = "path/to/metadata.rds"   # also accepts .csv or .tsv
)

```


# From a 10X Genomics output

The `create_sce()` function can also directly import from the standard 10X Genomics output directory, which contains the `matrix.mtx.gz`, `barcodes.tsv`, and `features.tsv` files:

```{r create-sce-files-from-10x}

sce <- create_sce(
  mtx_file   = "path/to/matrix.mtx.gz",
  cells_file = "path/to/barcodes.tsv",
  genes_file = "path/to/features.tsv"
)
```


# Next Steps

After creating the SCE, the next step is to perform quality control and filtering of low-quality cells, followed by normalisation of the expression data. These steps are described in more detail in the [Preprocessing and Analysis Tracks](preprocessing.html) article.
